Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-08-01 writeups/paper.tex
stash

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\bm{z} &=
            0\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\bm{u} \cdot \bm{\nabla}}$ is the Lagrangian or
material derivative, and $\bm{u}, \rho, P$ denote the velocity field, density
and pressure respectively. We denote $-g\bm{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\bm{\nabla}\overline{P} = -\overline{\rho} g\bm{z}$ and so $\overline{P} =
\overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\bm{x}$ is permitted in hydrostatic equilibrium, but we will
assume no background flow, and $\bm{u} = \bm{u}'$. Physically, this assumption
corresponds to a non-rotating star, or going to the corotating frame of a
rigidly rotating star.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ (e.g.\ see \citep{lecoanet_anel}). This both
identically enforces $\rho > 0$ and avoids numerical issues in the
$\frac{\bm{\nabla}P}{\rho}$ term if $\rho$ is small. We also define reduced
pressure $\varpi = \frac{P}{\rho}$. Then, we may rewrite the second two
equations of \autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u'_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}'}{t} + \bm{\nabla}\varpi
            + \varpi\bm{\nabla}\Upsilon + g\bm{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\Upsilon = 1, \overline{\varpi} = gH$.

@@

In the weak stratification limit $k_{z}H \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows with $z$ as $e^{z/2H}$. Thus, the
        linear approximation is always violated for sufficiently large $z$.

    \item The phase and group velocities are respectively:
        \begin{align}
            \bm{c}_{ph} &=
                \p{k_{x}\bm{x} + k_{z}\bm{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\bm{x}
                - \p{k_{x}k_{z}\bm{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{ph} \cdot \bm{c}_g =
        \mathcal{O}\p{\frac{1}{\p{k_{z}H}^{2}}} \approx 0$, the usual result
        (e.g.~\cite{drazin,sutherland1}). Note again that $c_{gz} < 0$ for
        $k_{x}, k_{z} > 0$, so upwards-propagating IGW have $k_z < 0$.

    \item The time-averaged total $x$-momentum flux $F$ carried in the $\bm{z}$
        direction can be computed. Since the linear solution consists of terms
        separable as $f(x, z, t) = f(z)e^{i(k_{1x}x - \omega_1 t)}$, $x$
        averaging and time averaging are equivalent. We may then write
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging. Note that indeed for an
        upwards propagating IGW $k_z < 0, k_x > 0$, we have $F > 0$.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. We implement forcing with strength $C$ by replacing
\autoref{eq:nl_up_density} with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}'\pd{\ln \overline{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}, \omega(k_{x}, k_{z})$ will propagate.

@@

This is in agreement with the breaking location predicted by the wave-induced
mean flow, given by \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{x}} \equiv \overline{U} = \frac{1}{L_x}
        \int\limits u_{x}\;\mathrm{d}x = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $F$ (\autoref{eq:F_def}) into a medium at
rest. Solving for $\overline{U} = \overline{U}_c \equiv \omega / k_x$ the
critical layer condition (see \autoref{ss:crit_layer} immediately below), we
obtain $\abs{\frac{u_{x}'k_{x}}{\omega}} = \abs{\xi_x k_{x}} = \abs{\xi_z k_z}
\sim 1$, equal to the nonlinearity criterion \autoref{eq:nl}. This condition
predicts where the wave's self-acceleration alone produces a critical layer.

@@

A purely horizontal shear flow $\overline{U}(z) \bm{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\overline{U}(z)\partial_x$. For a critical value $\omega - \overline{U}_c k_{x}
= 0$, the horizontal wavenumber of the linear wave in the fluid's frame of
reference vanishes and critical behavior is observed. In a linear
\citep{booker_bretherton} or viscous \citep{hazel} theory, the incident wave has
amplitude reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\overline{U}(z_c) = \frac{\omega}{k_{x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the mean flow cannot exceed $\overline{U}_c$ the
horizontal phase velocity of the incident wave, the critical layer must instead
propagate downwards to accommodate the incident momentum. The total horizontal
momentum of the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta S = 0.
\end{equation}
$\Delta S$ is the total injected horizontal momentum to the flow. We next make a
simple two-zone approximation $\overline{U}(z > z_c) = \overline{U}_c,
\overline{U}(z < z_c) = 0$, assuming all horizontal momentum flux goes into
critical layer propagation. Then, the conservation equation becomes simple
differential equation
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta S.\label{eq:zc_anal}
\end{equation}
For constant $\Delta S$ in time and $\rho \approx \overline{\rho}$, this has
analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\overline{\rho}_0\overline{U}_c}{-\Delta S}
        .\label{eq:zc_sol}
\end{equation}
Note that constant $-\Delta S$ corresponds physically to the picture where
constant $S$ carried by the IGWs, given by \autoref{eq:S_lin}, is fully absorbed
at the critical layer, in which case $\Delta S = -S$.

@@

We nondimensionalize by taking $H = N = \overline{\rho}_0 = 1$ in
\autoref{se:nl_orig}.

@@

    \caption{Snapshots of $u_x, u_z$ in the fiducial simulation illustrating
    distinct phases of the evolution of the flow. Note that $\bar{U}_c \approx
    0.16$ for the parameters used.}\label{fig:snapshots}
\end{figure*}

Slices of
$\overline{U}, \hat{F}$ across $z$ at various times $t$ are shown in
\autoref{fig:nl_fluxes}. While the behavior of $\overline{U}$ seems to conform
qualitatively with the predictions of \autoref{ss:crit_layer}, the behavior of
$\hat{F}$ exhibits two salient features: (i) the incident flux seems to
fluctuate greatly with time, and (ii) there seems to be a small transmitted
feature at many of the later times. We will discuss further these features in
\autoref{ss:reflectivity}, after first analyzing the propagation of the critical
layer.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/nl_fluxes.png}
    \caption{Plot of $\overline{U}(z, t)$ and $\hat{F}(z, t)$ respectively at
    various times over $z$ in our $\mathrm{Re} = 1024$ simulation.
    $\overline{U}, \hat{F}$ follow their definitions in \autoref{eq:mean_flow}
    and \autoref{eq:sbm_def}. $\overline{U}$ is given in units of $c_{ph, x} =
    \frac{\omega}{k_{x}}$ the critical layer flow velocity (see
    \autoref{ss:crit_layer}). The propagation of the critical layer towards
    lower $z$ and sharp deposition of $F$ at the critical layer are
    evident.}\label{fig:nl_fluxes}
\end{figure}

@@

The observation of reflected/transmitted waves is in accordange with previous
studies as discussed in \autoref{ss:wave_breaking}. Many studies have shown that
a local Richardson number $\mathrm{Ri} \sim 1/4$ corresponds to the onset of
reflectivity. This agrees with the observed Kelvin-Helmholtz instability in our
simulations [TODO image]. To verify that the critical layer's width is bound
from below by Kelvin-Helmholtz instabilities, we decided to compute a
$\mathrm{Ri}$ within the critical layer.

Since fluid instabilities are local, $\mathrm{Ri}$ must be measured in the
immediate vicinity of a fluid parcel. Thus, we first assign an $\mathrm{Ri}$ for
every $x$ in the critical layer, then take the median as $\mathrm{Ri}$ for the
entire layer. To avoid noisiness, the local $\mathrm{Ri}$ is computed using the
vertical distance over which the local $u_x$ increases from $0.3 \times$ its
critical value to its critical value. The value $0.3$ effectively excludes the
self-acceleration induced by the IGW\@. This can be written
\begin{align}
    z_{CL, \min}(x, t) &= \argmin_\zeta
        \z{z: u_x(x, \zeta, t) > 0.3\overline{U}_c},\nonumber\\
    z_{CL, \max}(x, t) &= \argmax
        \zeta \z{z: u_x(x, \zeta, t) < \overline{U}_c},\nonumber\\
    \mathrm{Ri}(t) &\equiv
        \med_x\p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}}.\label{ss:ri_med_def}
\end{align}
To understand the variation in $\mathrm{Ri}$ over $x$, we can also plot using
the minimum over $x$; the maximum is significantly noisier. Both of these are
shown in \autoref{fig:nl_f_ri}. The quick evolution of $\mathrm{Ri}$ to its
saturated value is reflective of the fact that IGW are \emph{anti-diffusive}.
This property is a simple consequence of the IGW dispersion relation
\autoref{eq:disp_rel}, which is approximately $\omega k_z \approx Nk_x$ such
that as an IGW propagates into a shear flow, $\omega$ decreases and $k_z$
increases, enhancing dissipation and further driving mean flow acceleration.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number of the flow at the critical layer over time
    as defined in \autoref{ss:ri_med_def}. Dotted red lines demarcate the
    minimum and maximum of $\mathrm{Ri}$ over $x$ instead of the median. These
    numbers effectively measure the mean and spread in width of the critical
    layer over $x$. Note that $\mathrm{Ri} \lesssim \frac{1}{4}$ corresponds
    to Kelvin-Helmholtz instability, so this plot suggests the critical layer
    width roughly saturates the lower bound from the Kelvin-Helmholtz
    instability after an initial transient phase.}\label{fig:nl_f_ri}
\end{figure}

@@

We identify two instances of non-absorptive behavior: (i), the presence of a
reflected wave with wave vector $\bm{k} = k_{x}\bm{x} + k_{z}\bm{z}$, and
(ii) the amount of horizontal momentum flux $S$ reflected off the critical
layer.

@@

To estimate convergence, we compute the median value of each of $\mathrm{Ri},
\mathcal{R}_A^2, \mathcal{R}_F, \mathcal{T}_F$ over the last $1/4$ of simulation
times from each simulation, where all simulations had converged to what appeared
to be asymptotic values. Error bars are estimated with the $16$ and $84$
percentiles. We illustrate the convergence of these averages across simulations
in \autoref{fig:agg}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/agg.png}
    \caption{Convergence of median reflection/transmission coefficients and
    $\mathrm{Ri}$ (critical layer width) across runs with varying viscosity,
    parameterized with $\mathrm{Re}$ (\autoref{eq:re_def}). Error bars depict
    $16$ and $84$ percentile values. Fmall horizontal displacements are made for
    data points at identical $\mathrm{Re}$ for readability. Note that
    simulations with larger $\mathrm{Re}$ correspond to smaller viscosity and
    are more physically realistic, and our values seem to converge towards large
    $\mathrm{Re}$.}\label{fig:agg}
\end{figure}

@@

\begin{itemize}
    \item $k_x, k_z$: In astrophysical systems, $k_{\perp} \ll k_r$. While we do
        not explore different $k_{1x}, k_{1z}, \omega_1$ in this study, with
        outgoing boundary conditions, there appears at first to be no $z$ length
        scale other than $k_{1z}$, so our results would seem to be invariant
        under rescaling of the $z$ length scale. However, true turbulence is
        expected to be isotropic at small scales, which may couple $k_{1x},
        k_{1z}$ in a way that $k_{1x} \ll k_{1z}$ produces different dynamics
        than $k_{1x} \lesssim k_{1z}$ as we've studied. This is a numerically
        difficult regime though, so we defer consideration to future work.

@@

Numerically, the nonlinear $\frac{\bm{\nabla}P}{\rho}$ term is problematic: we
desire a system where the fluid fields are not divided by one another. We
introduce $\varpi = \frac{P}{\rho}$ instead, then mandate $\overline{\rho}, \varpi_0$
background fields satisfy hydrostatic equilibrium $\bm{\nabla}\varpi_0 +
\varpi_0 \bm{\nabla}\overline{\rho} + g\bm{z} = 0$. Taking isothermal stratification,
we find $\varpi_0 = gH$. We further change variables to $\Upsilon = \ln \rho -
\ln \overline{\rho}$ and $\varpi_1 = \varpi - \varpi_0$ deviations from the background
state to obtain a system of equations at most quadratic in fluid fields:
\begin{subequations}\label{se:nl_var}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}_1 &= 0,\\
        \pd{\Upsilon}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}} \Upsilon
            - \frac{u_z}{H} &= 0,\\
        \pd{u_{1x}}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}}u_{1x}
            + \pd{\varpi_1}{x} + gH\pd{\Upsilon}{x}
            + \varpi_1 \pd{\Upsilon}{x} &= 0,\\
        \pd{u_z}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}}u_z
            + \pd{\varpi_1}{z} + gH\pd{\Upsilon}{z}
            + \varpi_1 \pd{\Upsilon}{z} - \frac{\varpi_1}{H} &= 0.
    \end{align}
\end{subequations}
It bears noting that these equations are exactly equivalent to the original
Euler equations and hence conserve horizontal momentum.

>>>>

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\bm{z} &=
            0\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\bm{u} \cdot \bm{\nabla}}$ is the Lagrangian or
material derivative, and $\bm{u}, \rho, P$ denote the velocity field, density
and pressure respectively. We denote $-g\bm{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\bm{\nabla}\overline{P} = -\overline{\rho} g\bm{z}$ and so $\overline{P} =
\overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\bm{x}$ is permitted at hydrostatic equilibrium, but we will
assume no background flow, so $\bm{u} = \bm{u}'$. Physically, this assumption
corresponds to a non-rotating star, or going to the corotating frame of a
rigidly rotating star.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ (e.g.\ see \citep{lecoanet_anel}). This both
identically enforces $\rho > 0$ and avoids numerical issues in the
$\frac{\bm{\nabla}P}{\rho}$ term if $\rho$ is small. We also define reduced
pressure $\varpi = \frac{P}{\rho}$. Then, we may rewrite the second two
equations of \autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u'_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}'}{t} + \bm{\nabla}\varpi
            + \varpi\bm{\nabla}\Upsilon + g\bm{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\Upsilon = 0, \overline{\varpi} = gH$.

@@

In the weak stratification limit $\abs{k_{z}H} \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows with $z$ as $e^{z/2H}$. Thus, the
        linear approximation is always violated for sufficiently large $z$.

    \item The phase and group velocities are respectively:
        \begin{align}
            \bm{c}_{ph} &=
                \p{k_{x}\bm{x} + k_{z}\bm{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\bm{x}
                - \p{k_{x}k_{z}\bm{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{ph} \cdot \bm{c}_g =
        \mathcal{O}\p{\frac{1}{\p{k_{z}H}^{2}}} \approx 0$, the usual result
        (e.g.~\cite{drazin,sutherland1}). Note that for an upwards propagating
        IGW $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

    \item The time-averaged total $x$-momentum flux $F$ carried in the $\bm{z}$
        direction can be computed. Since the linear solution consists of terms
        separable as $f(x, z, t) = f(z)e^{i(k_{x}x - \omega t)}$, $x$
        averaging and time averaging are equivalent. We may then write
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging. Note that indeed for an
        upwards propagating IGW $k_z < 0, k_x > 0$, we have $F > 0$.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. We implement forcing with strength $C$ localized around
$z_0$ with width $\sigma$ by replacing \autoref{eq:nl_up_density} with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}'\pd{\ln \overline{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}, \omega(k_{x}, k_{z})$ will propagate.

@@

This is in agreement with the breaking location predicted by the wave-induced
mean flow, given by \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{x}} \equiv \overline{U} = \frac{1}{L_x}
        \int\limits u_{x}\;\mathrm{d}x = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $F$ (\autoref{eq:F_def}) into a medium at
rest. Solving for $\overline{U} = \omega / k_x$ the critical layer condition
(\autoref{eq:u_c} and discussion in \autoref{ss:crit_layer}), we obtain
$\abs{\frac{u_{x}'k_{x}}{\omega}} = \abs{\xi_x k_{x}} = \abs{\xi_z k_z} \sim 1$,
equal to the nonlinearity criterion \autoref{eq:nl}. This condition predicts
where the wave's self-acceleration alone produces a critical layer.

@@

A purely horizontal shear flow $\overline{U}(z) \bm{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\overline{U}(z)\partial_x$. If anywhere $\overline{U}(z) = \overline{U}_c$ where
\begin{equation}
    \overline{U}_c \equiv \omega / k_x,\label{eq:u_c}
\end{equation}
then the horizontal wavenumber of the linear wave in the fluid frame vanishes
and critical behavior is observed. In a linear theory, both inviscid
\citep{booker_bretherton} and viscous \citep{hazel}, the incident wave has
amplitude reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R}_A &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T}_A &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\overline{U}(z_c) = \frac{\omega}{k_{x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the mean flow cannot exceed $\overline{U}_c$ the
horizontal phase velocity of the incident wave, the critical layer must instead
propagate downwards to accommodate the incident momentum. The total horizontal
momentum of the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta F = 0.
\end{equation}
$\Delta F$ is the total injected horizontal momentum to the flow. We next make a
simple two-zone approximation $\overline{U}(z > z_c) = \overline{U}_c,
\overline{U}(z < z_c) = 0$, assuming all horizontal momentum flux goes into
critical layer propagation. Then, the conservation equation becomes simple
differential equation
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta F.\label{eq:zc_anal}
\end{equation}
If the entirety of the horizontal momentum flux carried by the IGW wavetrain is
absorbed by the critical layer, then $\Delta F = -F'$ is constant in time.
For constant $\Delta F$ in time and $\rho \approx \overline{\rho}$, $z_c(t)$ has
analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\overline{\rho}_0\overline{U}_c}{-\Delta F}
        .\label{eq:zc_sol}
\end{equation}

@@

We nondimensionalize by taking $H = N = \overline{\rho}_0 = 1$.

@@

    \caption{Snapshots of $u_x, u_z$ in the fiducial simulation illustrating
    distinct phases of the evolution of the flow. Note that $\overline{U}_c \approx
    0.16$ for the parameters used.}\label{fig:snapshots}
\end{figure*}

Slices of
$\overline{U}, \hat{F}$ across $z$ at various times $t$ are shown in
\autoref{fig:nl_fluxes}. While the behavior of $\overline{U}$ seems to conform
qualitatively with the predictions of \autoref{ss:crit_layer}, the behavior of
$\hat{F}$ exhibits two salient features: (i) the incident flux seems to
fluctuate greatly with time, and (ii) there seems to be a small transmitted
feature at many of the later times. We will discuss further these features in
\autoref{ss:reflectivity}, after first analyzing the propagation of the critical
layer.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/nl_fluxes.png}
    \caption{Plot of $\overline{U}(z, t)$ and $\hat{F}(z, t)$ respectively at
    various times over $z$ in our $\mathrm{Re} = 1024$ simulation.
    $\overline{U}, \hat{F}$ follow their definitions in \autoref{eq:mean_flow}
    and \autoref{eq:sbm_def}. $\overline{U}$ is given in units of
    $\overline{U}_c = \frac{\omega}{k_{x}}$ the critical horizontal flow
    velocity (see \autoref{ss:crit_layer}). The propagation of the critical
    layer towards lower $z$ and sharp deposition of $F$ at the critical layer
    are evident.}\label{fig:nl_fluxes}
\end{figure}

@@

The observation of reflected/transmitted waves is in accordange with previous
studies as discussed in \autoref{ss:wave_breaking}. Studies have shown that a
local Richardson number $\mathrm{Ri} \sim 1/4$ corresponds to the onset of
reflectivity. This $\mathrm{Ri}$ also corresponds to the onset of the
Kelvin-Helmholtz instability (KHI). Indeed, in our simulations, visual
inspection suggests the KHI is present in the critical layer (see
\autoref{fig:snapshots}). It is natural to suspect then that the shear flow
cannot steepen any further than KHI onset. To verify this, we decided to compute
a local $\mathrm{Ri}$ for the shear flow around the critical layer.

Since fluid instabilities are local, $\mathrm{Ri}$ must be measured in the
immediate vicinity of a fluid parcel. Thus, we first assign an $\mathrm{Ri}$ for
every $x$ in the critical layer, then take the median as $\mathrm{Ri}$ for the
entire layer. To avoid noisiness, the local $\mathrm{Ri}$ is computed using the
vertical distance over which the local $u_x$ increases from $0.3 \times$ its
critical value to its critical value. The value $0.3$ effectively excludes the
self-acceleration induced by the IGW\@. This can be written
\begin{align}
    z_{CL, \min}(x, t) &= \argmin_\zeta
        \z{z: u_x(x, \zeta, t) > 0.3\overline{U}_c},\nonumber\\
    z_{CL, \max}(x, t) &= \argmax
        \zeta \z{z: u_x(x, \zeta, t) < \overline{U}_c},\nonumber\\
    \mathrm{Ri}(t) &\equiv
        \med_x\p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}}.\label{ss:ri_med_def}
\end{align}
To understand the variation in $\mathrm{Ri}$ over $x$, we can also plot using
the minimum over $x$ (the maximum is significantly noisier). Both of these are
shown in \autoref{fig:nl_f_ri}. The quick evolution of $\mathrm{Ri}$ to its
saturated value is reflective of the fact that IGW are \emph{anti-diffusive}.
This property is a simple consequence of the IGW dispersion relation
\autoref{eq:disp_rel}, which is approximately $\omega k_z \approx Nk_x$ such
that as an IGW propagates into a shear flow, $\omega$ decreases and $k_z$
increases, enhancing dissipation and further driving mean flow acceleration.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number of the flow at the critical layer over time
    as defined in \autoref{ss:ri_med_def}. Dotted red lines demarcate the
    minimum and median of $\mathrm{Ri}$ over $x$. These numbers effectively
    measure the mean and spread in width of the critical layer over $x$. Note
    that $\mathrm{Ri} \sim \frac{1}{4}$ corresponds to the KHI, so this plot
    suggests the shear at the critical layer does not steepen past the KHI
    onset.}\label{fig:nl_f_ri}
\end{figure}

@@

We identify two instances of non-absorptive behavior: (i), the presence of a
reflected wave with wave vector $\bm{k} = k_{x}\bm{x} + k_{z}\bm{z}$, and
(ii) the amount of horizontal momentum flux $F$ reflected off the critical
layer.

@@

To estimate convergence, we compute the median value of each of $\mathrm{Ri},
\mathcal{R}_A^2, \mathcal{R}_F, \mathcal{T}_F$ over the last $1/4$ of simulation
times from each simulation, where all simulations had converged to what appeared
to be asymptotic values. Error bars are estimated with the $16$ and $84$
percentiles. We illustrate the convergence of these averages across simulations
in \autoref{fig:agg}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/agg.png}
    \caption{Convergence of median reflection/transmission coefficients and
    $\mathrm{Ri}$ (critical layer width) across runs with varying viscosity,
    parameterized with $\mathrm{Re}$ (\autoref{eq:re_def}). Error bars depict
    $16$ and $84$ percentile values. Small horizontal displacements are made for
    data points at identical $\mathrm{Re}$ for readability. Note that
    simulations with larger $\mathrm{Re}$ correspond to smaller viscosity and
    are more physically realistic, and our values seem to converge towards large
    $\mathrm{Re}$. At the smallest $\mathrm{Re}$ value, $\mathrm{Ri} \approx 50$
    is too large to fit on the plot.}\label{fig:agg}
\end{figure}

It is apparent that the Richardson number rapidly converges to $\mathrm{Ri}
\approx 0.4$ but the reflection/transmission coefficients converge more slowly.
This is in disagreement with \autoref{eq:crit_coeffs} calculated via the linear
analytical theory. This tension is natural: fluid motion within the critical
layer is turbulent, so transmission/reflection at the critical layer cannot be
captured by any linear theory.

@@

\begin{itemize}
    \item $k_x, k_z$: In astrophysical systems, $k_{\perp} \ll k_r$. While we do
        not explore different $k_{x}, k_{z}, \omega_1$ in this study, with
        outgoing boundary conditions, there appears at first to be no $z$ length
        scale other than $k_{z}$, so our results would seem to be invariant
        under rescaling of the $z$ length scale. However, true turbulence is
        expected to be isotropic at small scales, which may couple $k_{x},
        k_{z}$ in a way that $k_{x} \ll k_{z}$ produces different dynamics
        than $k_{x} \lesssim k_{z}$ as we've studied. This is a numerically
        difficult regime though, so we defer consideration to future work.

@@

Numerically, the nonlinear $\frac{\bm{\nabla}P}{\rho}$ term is problematic: we
desire a system where the fluid fields are not divided by one another. We
introduce $\varpi = \frac{P}{\rho}$ instead, then mandate $\overline{\rho}, \varpi_0$
background fields satisfy hydrostatic equilibrium $\bm{\nabla}\varpi_0 +
\varpi_0 \bm{\nabla}\overline{\rho} + g\bm{z} = 0$. Taking isothermal stratification,
we find $\varpi_0 = gH$. We further change variables to $\Upsilon = \ln \rho -
\ln \overline{\rho}$ and $\varpi_1 = \varpi - \varpi_0$ deviations from the background
state to obtain a system of equations at most quadratic in fluid fields:
\begin{subequations}\label{se:nl_var}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}_1 &= 0,\\
        \pd{\Upsilon}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}} \Upsilon
            - \frac{u_z}{H} &= 0,\\
        \pd{u_{x}}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}}u_{x}
            + \pd{\varpi_1}{x} + gH\pd{\Upsilon}{x}
            + \varpi_1 \pd{\Upsilon}{x} &= 0,\\
        \pd{u_z}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}}u_z
            + \pd{\varpi_1}{z} + gH\pd{\Upsilon}{z}
            + \varpi_1 \pd{\Upsilon}{z} - \frac{\varpi_1}{H} &= 0.
    \end{align}
\end{subequations}
It bears noting that these equations are exactly equivalent to the original
Euler equations and hence conserve horizontal momentum.
